Monocle analysis of influenza-infected cells

### 2017-06-04 monocle_analysis.Rmd
Plot total mRNAs for infected and un-infected cells

## Create and filter cell-gene dataset
Read in cell-gene data of all samples.
This is the merged cell-gene data set constructed essentially from `cellranger` using the iPython notebook.
The cells are annotated by the number of wildtype and synonymous-barcoded influenza UMIs called for each cell.
```{r}
matrixfile <- "results/cellgenecounts/merged_humanplusflu_matrix.mtx"
genesfile <- "results/cellgenecounts/merged_humanplusflu_genes.tsv"
cellsfile <- "results/cellgenecounts/merged_humanplusflu_cells.tsv"

@@

Filter on just the samples that interest us for this analysis.
```{r}
samples = c("Uninfected", "8hr-2", "8hr-2repeat", "10hr-2")

@@

Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
```{r}
pData(cells)$total_mRNAs <- Matrix::colSums(exprs(cells))

flugenes = c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
flu_id = row.names(subset(fData(cells), gene_short_name %in% flugenes))
pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flu_id,]))

flugenes_id = row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))
```

```{r}
scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()

@@

scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5)) + facet_wrap(~Sample)
```
The scatterplots above show the number of influenza mRNAs versus the number of cellular mRNAs for each sample. 
The blue curves show the overall density on the number of cellular mRNAs.
Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
I have **not** done that here for the reason that this will filter out quite a few of the influenza-infected cells.
This fact is made more clear by this next plot:
```{r}
all_densityplot <- ggplot(pData(cells), aes(total_mRNAs)) + geom_density()
flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 100))
flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
flugt100_densityplot + facet_wrap(~Sample)
```

>>>>

## Create cell-gene dataset and examine some basic statistics about the number of influenza and cellular reads per cell
Read in cell-gene data of all samples.
This is the merged cell-gene data set constructed essentially from `cellranger` using the iPython notebook.
The cells are annotated by the number of wildtype and synonymous-barcoded influenza UMIs called for each cell.
```{r}
matrixfile <- "results/cellgenecounts/merged_humanplusflu_matrix.mtx"
genesfile <- "results/cellgenecounts/merged_humanplusflu_genes.tsv"
cellsfile <- "results/cellgenecounts/merged_humanplusflu_cells.tsv"

@@

Filter on just the samples that interest us for this analysis.
Right now this is all the samples in the cell-gene dataset.
```{r}
samples = c("Uninfected", "6hr-1", "8hr-2", "8hr-2repeat", "10hr-2")

@@

Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
Also compute the fraction of total mRNAs that are from influenza.
```{r}
pData(cells)$total_mRNAs <- Matrix::colSums(exprs(cells))

flugenes <- c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
flugenes_id <- row.names(subset(fData(cells), gene_short_name %in% flugenes))
pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))

pData(cells)$frac_flu_mRNAs <- pData(cells)$flu_mRNAs / pData(cells)$total_mRNAs

cellgenes_id <- row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[cellgenes_id,]))
```

Plot the number of influenza mRNAs versus the number of cellular RNAs as scatterplots.
The blue curves show the overall density on the number of cellular mRNAs.
Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
```{r}
scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()

@@

scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5)) + facet_wrap(~Sample)
```

The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
I have **not** done that for the reason that this will filter out quite a few of the influenza-infected cells.
This fact is made more clear by this next plot, which shows that the cells that have greater than 1% of their reads derived from influenza tend to have more total mRNAs than other cells in the two samples with the greatest number of influenza reads
I can think of several possible reasons for this trend: 
  1. Cells infected with influenza are simply "bigger" or are stimulated to produce more mRNAs
  2. Influenza infection preferentially leads to doublets due to cell clumping, perhaps mediated by hemagglutinin-based physical linkage of cells.
In any case, it seems safer **not** to exclude the influenza infected cells as we'd then be preferentially excluded cells with high viral burden.
```{r}
pData(cells)$frac_flu_gt_0.01 <- pData(cells)$frac_flu_mRNAs > 0.01

ggplot(pData(cells[, row.names(subset(pData(cells), Sample %in% c("8hr-2", "10hr-2")))]), 
        aes(total_mRNAs)) +
    geom_density(aes(color=frac_flu_gt_0.01)) +
    theme(axis.text.x=element_text(angle=90, vjust=0.5)) +
    facet_wrap(~Sample)
#flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 50))
#flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
```
